---
title: "scRNA-seq analysis using Seurat package"
output: html_notebook
author: Long Nguyen
---
# Description:
* The purpose of this notebook is to analyze scRNA-seq data of mouse Estrus
* The method used is reading directly from mtx files (CellRanger count outputs)
* The algorithms used come from Seurat package
* Two mitochondiral genes: mt-Nd6 and mt-Cytb will be manually deleted
* Because read_loom method in Scanpy does not contain those genes after filtering was done
* The results will compared to:
  + Same dataset analyzed using Scanpy - ManuallyRemoveGenes
  + Same dataset analyzed using Scanpy - readloom
  
# Workflow begin
This tutorial comes from [Seurat](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)

## Load required packages
```{r}
library(dplyr)
library(Seurat)
library(leiden)
library(ggplot2)
library(patchwork)
```


## Load data
Input is a directory that contains CellRanger count outputs
```{r}
previous_time <- proc.time()[3]
Estrus.data <-Read10X(data.dir ="/home/nguyen/Estrus_cellranger_count/Estrus_Raw/")
proc.time()[3] - previous_time
```
* View(Estrus.data) - run this command to investigate the output
* Current dataset:
  + 31253 genes
  + 6599 cells

## Initialize the Seurat object
* Initialize the Seurat object with the raw (non-normalized data).
* Filter out genes that are detected in less than 3 cells 
* Filter out cells that have less than 200 genes expressed
```{r}
previous_time <- proc.time()[3]
Estrus <- CreateSeuratObject(counts = Estrus.data, project = "Estrus", min.cells = 3, min.features = 200)
proc.time()[3] - previous_time
Estrus
```
* Output:
  + 19420 genes 
  + 6481 cells

## Mitochondiral genes overview
```{r}
# This is the list of genes that start with mt
mito_genes <- grep("^mt-", rownames(Estrus) , ignore.case=T, value=T)
print(mito_genes)
```

## mt-Nd6 and mt-Cytb will be manually deleted
```{r}
counts <- GetAssayData(Estrus, assay = "RNA")
counts <- counts[-(which(rownames(counts) %in% c("mt-Nd6", "mt-Cytb"))),]
Estrus <- subset(Estrus, features = rownames(counts))
```

```{r}
# This is the list of genes that start with mt
mito_genes <- grep("^mt-", rownames(Estrus) , ignore.case=T, value=T)
print(mito_genes)
```
* "mt-Nd6", "mt-Cytb" were deleted

## Calculate the percentage of mitochondrial genes
* Add "percent.mt" column to object metadata
* Percentage of mitochondrial genes is stored in "percent.mt"
* "^mt-" means strings start with mt (lowercase mt for mice, uppercase MT for human)
```{r}
Estrus[["percent.mt"]] <- PercentageFeatureSet(Estrus, pattern = "^mt-")
```

```{r}
# This is the data in "percent.mt" column
percent_mito <- FetchData(object = Estrus, vars = "percent.mt") #how to fetch data
head(percent_mito)
```

```{r, fig.width=10}
# Visualize QC metrics as a violin plot
VlnPlot(Estrus, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
```{r}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.
plot1 <- FeatureScatter(Estrus, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(Estrus, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

## Filter out low-quality cells
* We filter out cells that have unique feature counts over 6000
* Because any cell that has more than 6000 genes are likely doublets
* Note: Seurat tutorial used 2500 instead of 6000
* We also filter out cells that have more than 5% mitochondrial genes
* Because those are likely to be nonviable cells
```{r}
previous_time <- proc.time()[3]
Estrus <- subset(Estrus, subset = nFeature_RNA < 6000 & percent.mt < 5)
proc.time()[3] - previous_time
```
```{r}
Estrus
```
* After filtering step, the dataset is left with
 + 19418 genes
 + 2585 cells

## Normalizing the data
* Perform log-normalization method
* Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor
* scale.factor default value = 10,000
* This is then natural-log transformed using log1p
```{r}
previous_time <- proc.time()[3]
Estrus <- NormalizeData(Estrus, normalization.method = "LogNormalize", scale.factor = 10000)
proc.time()[3] - previous_time
```
## Identification of highly variable features (feature selection)
```{r}
previous_time <- proc.time()[3]
Estrus <- FindVariableFeatures(Estrus, selection.method = "disp", nfeatures = 5000)
proc.time()[3] - previous_time
```
## Plotting highly variable features
```{r}
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(Estrus), 10)
top10
```

```{r, fig.width=10, fig.height=4}
# Plot variable features with and without labels
plot1 <- VariableFeaturePlot(Estrus)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2 
```
## Regress out data and perform scaling
* Regressing out total counts per cell and percent.mt before scaling. 
* Take out mitochondrial genes before scaling (analyzing) the data 
* Because mito_genes shouldn't be accounted for cell population analysis.
```{r}
previous_time <- proc.time()[3]
all.genes <- rownames(Estrus)
Estrus <- ScaleData(Estrus, features = all.genes, vars.to.regress = c("nCount_RNA","percent.mt"), scale.max = 10)
# Explanations below are from [Seurat](https://satijalab.org/seurat/archive/v3.0/cell_cycle_vignette.html)
# vars.to.regress = RegressOut in Seurat version 1.4 or lower
# However, as the results of this procedure are stored in the scaled data slot (therefore overwriting the output of ScaleData) 
# This functionality is merged into the ScaleData function itself.
proc.time()[3] - previous_time
```
## Perform linear dimensional reduction (PCA)
```{r}
previous_time <- proc.time()[3]
Estrus <- RunPCA(Estrus, features = VariableFeatures(object = Estrus))
proc.time()[3] - previous_time
```

## Examine and visualize PCA results a few different ways
```{r}
print(Estrus[["pca"]], dims = 1:5, nfeatures = 5)
```

```{r}
VizDimLoadings(Estrus, dims = 1:2, reduction = "pca")
```

```{r}
DimPlot(Estrus, reduction = "pca") + ggtitle(label = 'Including immune cells') 
ggsave(filename = "PCA_Including_immune_cells.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```

```{r, fig.width=12}
FeaturePlot(Estrus, reduction = "pca", features = c('Ovgp1', 
                                                    'Foxj1', 'Igf1', 'Pdxk', 
                                                    'Fxyd4', 'Bsg', 
                                                    'Serpina1e', 'Krt19'))
ggsave(filename = "PCA_Including_immune_cells1.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
```{r, fig.width=12}
FeaturePlot(Estrus, reduction = "pca", features = c('Dcn', 'Pax8',
                                                    'Ephx2', 
                                                    'Pecam1', 'Pdgfra', 'Myh11', 'Atp2b4', 
                                                    'Cd52'))
ggsave(filename = "PCA_Including_immune_cells2.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```

## Determine the ‘dimensionality’ of the data
* Seurat tutorial uses JackStraw() and and ScoreJackStraw to determine dimensionality
* the syntax is: 
  + Estrus <- JackStraw(Estrus, num.replicate = 100)
  + Estrus <- ScoreJackStraw(Estrus, dims = 1:20)
* However, it takes very long. 
* An elbow plot often corresponds well enough with the significant dims and is much faster to run than Jackstraw
* According to the graph below, choose the first 22 principle components to use
```{r}
ElbowPlot(Estrus, ndims = 40)
```
## Find neighbors
* Dims = dimensions = determined as 22 from the above step
```{r}
previous_time <- proc.time()[3]
Estrus_new <- FindNeighbors(Estrus, dims = 1:22, k.param = 10)
proc.time()[3] - previous_time
```
## Find cell clusters using Leiden's method
```{r}
# Algorithm = 4 specifies Leiden's method
previous_time <- proc.time()[3]
Estrus_new <- FindClusters(Estrus_new, resolution = 0.14, algorithm = 4)
proc.time()[3] - previous_time
```
## Constructing UMAP plot
```{r}
previous_time <- proc.time()[3]
Estrus_new <- RunUMAP(Estrus_new, dim = 1:22) 
proc.time()[3] - previous_time
```
## Visualization
```{r}
DimPlot(Estrus_new, reduction = "umap") + ggtitle(label = 'Including immune cells')
ggsave(filename = "UMAP_Including_immune_cells.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
```{r, fig.width=12}
FeaturePlot(Estrus_new, reduction = "umap", features = c('Ovgp1', 'Foxj1', 'Dcn', 'Serpina1e', 'Sprr2f',
                                                         'Fxyd4', 'Ly6a', 'Epcam', 'Krt19', 'Vim', 'Pdgfra'))
ggsave(filename = "UMAP_Including_immune_cells1.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```

```{r, fig.width=12}
FeaturePlot(Estrus_new, reduction = "umap", features = c('Myh11', 
                                        'Dkk2', 'Slitrk6', 'Dio2', 'Smoc2', 'Tes', 'Dab1', 'Camk1d', 'Pecam1', 'Cd52','Mki67'))
ggsave(filename = "UMAP_Including_immune_cells2.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
## Removing immune cells
```{r, fig.width=12}
sub_Estrus = subset(x = Estrus_new, ident = c(1,2,3,4,5,6,7,9,10,12))
DimPlot(sub_Estrus, reduction = "umap") + ggtitle(label = 'Immune cells removed')
ggsave(filename = "UMAP_Immune_cells_removed.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
```{r, fig.width=12}
FeaturePlot(sub_Estrus, reduction = "umap", features = c('Ovgp1', 'Foxj1', 'Fxyd4', 'Cxcl17', 'Myh11', 'Cd52', 'Twist2', 'Pecam1'))
ggsave(filename = "UMAP_Immune_cells_removed1.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
```{r, fig.width=12}
FeaturePlot(sub_Estrus, reduction = "umap", features = c('Pdgfra', 'Ephx2', 'Dcn', 'Krt19', 'Epcam', 'Vim','Mki67'))
ggsave(filename = "UMAP_Immune_cells_removed2.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
```{r}
sub_Estrus
```
* After removing immune cells, the dataset is left with:
  + 19418 genes
  + 2318 cells

## Re-run analysis after excluding immune cells
```{r}
previous_time <- proc.time()[3]
sub_Estrus <- RunPCA(sub_Estrus, features = VariableFeatures(object = sub_Estrus))
proc.time()[3] - previous_time
```

```{r, fig.width=12}
DimPlot(sub_Estrus, reduction = "pca") + ggtitle(label = 'Immune cells removed')
ggsave(filename = "PCA_Immune_cells_removed.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```

```{r, fig.width=12}
FeaturePlot(sub_Estrus, reduction = "pca", features = c('Esr1', 'Ovgp1', 'Foxj1', 'Fxyd4', 'Ephx2', 'Dcn'))
ggsave(filename = "PCA_Immune_cells_removed1.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```

```{r, fig.width=12}
FeaturePlot(sub_Estrus, reduction = "pca", features = c('Fos', 'Kcne3', 'Serpina1e', 'Crabp2', 'Pgr', 'Vim'))
ggsave(filename = "PCA_Immune_cells_removed2.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
```{r}
previous_time <- proc.time()[3]
sub_Estrus_new <- FindNeighbors(sub_Estrus, dims = 1:22)
proc.time()[3] - previous_time
```

```{r}
previous_time <- proc.time()[3]
sub_Estrus_new <- FindClusters(sub_Estrus_new, resolution = 0.04, algorithm = 4)
proc.time()[3] - previous_time
```

```{r}
previous_time <- proc.time()[3]
sub_Estrus_new <- RunUMAP(sub_Estrus_new, dim = 1:22)
proc.time()[3] - previous_time
```

```{r, fig.width=10}
DimPlot(sub_Estrus_new, reduction = "umap") + ggtitle(label = 'Rescale UMAP after removing Immune cells')
ggsave(filename = "UMAP_Rescale.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
```{r, fig.width=12}
FeaturePlot(sub_Estrus_new, reduction = "umap", features = c('Ovgp1', 'Foxj1', 'Crabp2', 'Myh11', 'Wt1', 'S100g', 'Sprr2f',
                  'Fxyd4'))
ggsave(filename = "UMAP_Rescale1.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```

```{r, fig.width=12}
FeaturePlot(sub_Estrus_new, reduction = "umap", features = c('Dcn', 'Pax8', 'Pax2', 'Twist2', 'Igf1', 'Serpina1e', 'Pdgfra', 'Krt19', 'Mki67'))
ggsave(filename = "UMAP_Rescale2.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
## Marker genes
```{r}
marker_genes = c('Epcam', 'Krt15','Krt8','Krt19','Vim','Pdgfra','Twist2', 'Sprr2f', 'Fxyd4', 'Igfbp6',
                 'Acta2', 'Myh11','Foxj1','Ovgp1','Wt1','Serpina1e','S100g','Pecam1')
```

## UMAP for marker genes
```{r, fig.width=12}
FeaturePlot(sub_Estrus_new, reduction = "umap", features = c('Epcam', 'Krt15','Krt8','Krt19','Vim','Pdgfra','Twist2', 'Sprr2f', 'Fxyd4'))
ggsave(filename = "UMAP_for_Marker_genes1.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```

```{r, fig.width=12}
FeaturePlot(sub_Estrus_new, reduction = "umap", features = c('Igfbp6',
                                                             'Acta2', 'Myh11','Foxj1','Ovgp1','Wt1','Serpina1e','S100g','Pecam1'))
ggsave(filename = "UMAP_for_Marker_genes2.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```

## Rename clusters
```{r}
Cluster.names <- c('1 Ciliated_InfAmp',
                   '2 Secretory_InfAmp',
                   '3 Epithelial_subset',
                   '4 Fibroblast Pdgfra+',
                   '5 Secretory_IsthUTJ',
                   '6 Fibroblast Pdgfra-',
                   '7 Muscle')
names(Cluster.names) <- levels(sub_Estrus_new)
sub_Estrus_new <- RenameIdents(sub_Estrus_new, Cluster.names)
DimPlot(sub_Estrus_new, reduction = "umap", pt.size = 0.5) + scale_color_manual(labels = Cluster.names,values=c('green',                                                                                        'red','blue','orange','grey','yellow','pink','purple'))
ggsave(filename = "UMAP_Clusters_names.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
```{r, fig.width=12}
DotPlot(sub_Estrus_new, features = marker_genes, group.by = sub_Estrus_new@meta.data$ident) +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_y_discrete(labels = Cluster.names)
ggsave(filename = "DotPlot_for_Marker_genes.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
## Number of cells in each clusters
```{r}
table(sub_Estrus_new@active.ident, sub_Estrus_new@meta.data$orig.ident)
```

## Extract top genes for each cluster
* Default method of finding highly expressed genes for FindAllMarkers is Wilcoxon Rank Sum test
```{r}
previous_time <- proc.time()[3]
sub_Estrus_new_markers <- FindAllMarkers(sub_Estrus_new, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
sub_Estrus_new_markers %>%
    group_by(cluster) %>%
    slice_max(n = 20, order_by = avg_log2FC)
proc.time()[3] - previous_time
```
## Heat map for top 10 genes in each cluster
```{r, fig.width=10}
sub_Estrus_new_markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
DoHeatmap(sub_Estrus_new, features = top10$gene) + NoLegend()
ggsave(filename = "HeatMap_Top10_genes.pdf", plot = last_plot(), device = "pdf",
         path = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Figures/")
```
```{r}
saveRDS(sub_Estrus_new, file = "/home/nguyen/FinalWork/R/ManuallyRemoveGenes/Estrus_subset_analyzed.rds")
```

